---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r setup, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load necessary packages
library(ggthemes)
library(tidyverse)

# Code for reading in the dataset and/or processing the data

if(!file.exists("activity.csv")){
        unzip("activity.zip")}

activity <- read.csv("activity.csv")

activity$date <- as.Date(activity$date)

```


## What is mean total number of steps taken per day?
```{r mean steps per day}
# Calculate the totla number of steps taken per day
dailySteps <- activity %>%
        group_by(date) %>%
        summarise(totalSteps = sum(steps, na.rm = TRUE)) # ignore missing values

# Make a histogram of the total number of steps taken each day.
ggplot(dailySteps, aes(x = totalSteps)) +
        geom_histogram(bins = 20) +
        labs(title = "Histogram of the total number of steps taken daily",
             x = "Number of Steps",
             y = "Count") +
        theme_tufte()

# Calculate and report the mean and median of the total number of steps taken per day.
meanDailySteps <- mean(dailySteps$totalSteps)
medianDailySteps <- median(dailySteps$totalSteps)

```

The mean number of steps taken daily is `r round(meanDailySteps)`.

The median number of steps taken daily is `r medianDailySteps`.


## What is the average daily activity pattern?
```{r}
# Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) 
# and the average number of steps taken, averaged across all days (y-axis)
intervalAverageSteps <- activity %>%
        group_by(interval) %>%
        summarise(averageSteps = mean(steps, na.rm = TRUE))

ggplot(intervalAverageSteps, aes(x = interval, y = averageSteps)) +
        geom_line() +
        labs(title = "Average steps taken across 5-minute intervals",
             x = "Minutes",
             y = "Average Number of Steps") + 
        theme_tufte()

# Which 5-minute interval, on average across all the days in the dataset, 
# contains the maximum number of steps?
intervalMaxSteps <- with(intervalAverageSteps, 
                         interval[averageSteps == max(averageSteps)])

```

Time interval number `r intervalMaxSteps` contains the maximum number of steps, on average.

## Imputing missing values
```{r}
# Calculate and report the total number of missing values in the dataset.
missingCount <- sum(!complete.cases(activity))

```

The dataset contains `r missingCount` rows with missing data.

```{r}
# Strategy for imputing missing values: 
# replace NAs using the mean value for same interval across all days

# Create a new dataset that is equal to the original dataset but with the missing data filled in.
activityImputed <- activity %>%
        group_by(interval) %>%
        mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = TRUE), steps))

imputedDailySteps <- activityImputed %>%
        group_by(date) %>%
        summarise(totalSteps = sum(steps))

# Make a histogram of the total number of steps taken each day.
ggplot(imputedDailySteps, aes(x = totalSteps)) +
        geom_histogram(bins = 20) +
        labs(title = "Histogram of the total number of steps taken daily (missing values imputed)",
             x = "Number of Steps",
             y = "Count") +
        theme_tufte()

# Calculate and report the mean and median total number of steps taken per day.
meanImputedDailySteps <- mean(imputedDailySteps$totalSteps)
medianImputedDailySteps <- median(imputedDailySteps$totalSteps)

options(scipen = 999) # Turns off scientific notation format on displayed mean and median

```

By imputing the missing step counts, the mean daily step count increases from `r round(meanDailySteps)` to `r round(meanImputedDailySteps)`. The median value decreases from `r round(medianDailySteps)` to `r round(medianImputedDailySteps)`. With the imputed missing values, the mean and median step counts are equal. 

## Are there differences in activity patterns between weekdays and weekends?
```{r}
# Create a new factor variable in the dataset with two levels – “weekday” and “weekend”
# Make a panel plot of the 5-minute interval (x-axis) and the average number of steps taken, 
# averaged across all weekday days or weekend days (y-axis)
activityImputedDayType <- activityImputed %>%
        mutate(dayType = factor(ifelse(weekdays(date) %in% c("Saturday", "Sunday"), 
                                       "weekend", 
                                       "weekday"))) %>%
        group_by(dayType, interval) %>%
        summarise(averageSteps = mean(steps))

ggplot(activityImputedDayType, aes(x = interval, y = averageSteps)) +
        geom_line() +
        facet_grid(dayType~.) +
        labs(title = "Average steps taken across 5-minute intervals by type of day",
             x = "Minutes",
             y = "Average Number of Steps") + 
        theme_tufte()

```
